Code walkthrough
===

The code used for all examples in my Google DevFest presentation are contained in this file, along with comments to help you get started. Start off by installing/loading the requisite packages:

```{r}
install.packages("ggmap")
install.packages("maps")
install.packages("geosphere")
install.packages("rgdal")
install.packages("maptools")
install.packages("ggplot2")
install.packages("plyr")
install.packages("jsonlite")

library(ggmap)
library(maps)
library(geosphere)
library(rgdal)
library(maptools)
library(ggplot2)
library(plyr)
library(jsonlite)
```

# Translating string location to lat/lon

First up, to get latitude and longitude values for a location, just type what you wold type into Google Maps via the `geocode()` function in the `ggmap` package:


```{r }
  geocode("St. Paul, MN")

  geocode("2115 Summit Ave., St. Paul, MN")

  geocode("University of St. Thomas, MN")
```

# Getting maps

You can grab map images via a 1) center location/zoom level or 2) bounding box (min/mas lat and lon). You can pull from four different source: Google, Stamen, Cloudmade, and OpenStreeMaps.

Get a lat/lon to use as the center, then grab a satellite map from Google.

```{r }
loc <- geocode("2115 Summit Ave, St. Paul, MN")
ust <- get_map(location = c(lon = loc$lon, lat = loc$lat),
                zoom = 15, source = "google",
                maptype = "hybrid", crop = T)

# plot the map you grabbed
ggmap(ust)
```

If we specify borders, we can get a non-square map:

```{r }
# same center location
loc <- geocode("2115 Summit Ave, St. Paul, MN")

# now we calculate our borders with +/- lat and +/- lon values
box <- c(left = loc$lon - 0.04, bottom = loc$lat - 0.02,
         right = loc$lon + 0.04, top = loc$lat + 0.02)

# pulling from stamen this time with a nice watercolor style
ust_box <- get_map(location = box, source = "stamen",
                   maptype = "watercolor", crop = T)

ggmap(ust_box)
```

# Overplotting maps

-   `ggmap()` adds onto existing `ggplot2` functionality
-   Thus, layering standard `ggplot2` graphics on top of maps is easy!

```{r }
locs <- data.frame(names = c("st. paul", "minneapolis"))                  # grab locations
locs <- cbind(locs, geocode(as.character(locs$names)))

mid <- get_map(location = c(lon = mean(locs$lon), lat = mean(locs$lat)),  # get a map
               zoom = 10, source = "stamen", maptype = "toner", crop = T)

p <- ggmap(mid)                                                           # plot map
p <- p + geom_point(aes(x = lon, y = lat, colour = factor(names)),        # overplot w/ points
                    dat = locs, size = 6)
p <- p + scale_colour_discrete("Location")
```

# Working with world/country maps

If you need bigger data, `get_map()` won't work since the sources only have a certain allowable zoom level.

```{r }
# get map of the world
world <- map_data("world")

# get a sense for what that data looks like
head(map, 5)
tail(map, 5)

# plotting world map with ggplot2
# pay attention to column names (long vs. lon!)
p <- ggplot(world, aes(x = long, y = lat, group = group))
p <- p + geom_polygon(colour = "white")
p
```

There's also a `state` data set to plot the US with state borders.

```{r usa-map}
usa <- map_data("state")
p <- ggplot(usa, aes(x = long, y = lat, group = group))
p <- p + geom_polygon(colour = "white")
p
```

Perhaps you want to only plot a subset of states?

```{r some-states}
# make vector of desired states
states <- c("minnesota", "wisconsin", "illinois", "indiana",
            "iowa", "missouri", "michigan")

# get the full state map data
states_map <- map_data("state")

# use R's subsetting abilities to keep only what's in our list above
states_map <- states_map[states_map$region %in% states, ]

# plot
p <- ggplot(states_map, aes(x = long, y = lat, group = group))
p <- p + geom_polygon(colour = "white")
p
```


# Example: great circle data

3M has internal technology "clubs" (called chapters), which organize talks around various technologies. Typically these are given in the US and recorded for later viewing by OUS colleagues. We got a request to do this in reverse instead: give a talk at 6am Central Time in order to make it more appealing for OUS viewers and record for the US.

The belief was that a live talk would gather far more attendees, and the belief proved true! We did two talks like this and they shattered attendance records! I wanted to make an infographic to help encourage other chapters to use this method, and got the attendance record data (country and # viewers) and did it!

To start, I found the `gcIntermediate()` function from  `geosphere` package, which will calcualte a great circle arc between two locations. The syntax is fairly simple:

```{r eval=FALSE}
library(geosphere)
arc <- gcIntermediate(c(lon_1, lat_1), c(lon_2, lat_2),
                      n = steps, addStartEnd = T)
```

If you have points that cross the Date Line, R will return two sets of arc points instead of one. We have to catch this and put the two sets in a format compatible with plotting later. This is that part of the code:

```{r }
# draw great circles from St. Paul to everywhere else
gcircles <- lapply(1:nrow(end), function(i) {                 # arc for each set
  temp <- gcIntermediate(start[, c("lon", "lat")],
                         end[i, c("lon", "lat")],
                         n = 50, addStartEnd = T,
                         breakAtDateLine = T)

  if(is.list(temp) == T) {                                    # were two half-paths returned?
    ids <- c(rep(paste0("i", i), nrow(temp[[1]])),            # if so, make id's for both halves
             rep(paste0("j", i), nrow(temp[[2]])))
    temp <- as.data.frame(rbind(temp[[1]], temp[[2]]))        # combine two sets of points
    temp$id <- ids                                            # assign ids for plotting
  }
  ...
```


Here's the full plot to create a map with points sized according to attendees and great circle arcs connecting St. Paul to each location.

```{r great-circles}
# load data
talks <- read.csv("./data/talk-locations.csv")

# total live viewers as well as those who watched the recording later
talks$total <- talks$live + talks$rec

# there were two talks, so we join each city/country's records from both talks
# into a single row per location.
talks_agg <- ddply(talks, .(country, city), summarize,
                   total = sum(total), lat = lat[1], lon = lon[1])

# for the arcs, the start will always be St. Paul so we split that out
start <- talks_agg[talks_agg$city == "Saint Paul", ]
end <- talks_agg[talks_agg$city != "Saint Paul", ]

# draw great circles from St. Paul to everywhere else
gcircles <- lapply(1:nrow(end), function(i) {
  temp <- gcIntermediate(start[, c("lon", "lat")], end[i, c("lon", "lat")],
                         n = 50, addStartEnd = T, breakAtDateLine = T)

  # if temp is a list, rbind and return single list
  if(is.list(temp) == T) {
    ids <- c(rep(paste0("i", i), nrow(temp[[1]])),
             rep(paste0("j", i), nrow(temp[[2]])))
    temp <- as.data.frame(rbind(temp[[1]], temp[[2]]))
    temp$id <- ids
  }

  else {
    ids <- rep(paste0("i", i), nrow(temp))
    temp <- as.data.frame(temp)
    temp$id <- ids
  }
  
  return(temp)
  
})

# lapply returns a list of the results; combine into a single table
gcircles <- do.call(rbind, gcircles)

# get map data and remove the way south stuff we don't need
world <- map_data("world")
world <- world[world$lat > -57, ]

# initialize a ggplot object; need to do it this way since
# we're using independent layers that don't share same aes params
p <- ggplot()

# plot the world map
p <- p + geom_polygon(aes(x = long, y = lat, group = group),
                      data = world, colour = "gray10",
                      fill = "gray95")

# plot the great circles
p <- p + geom_line(aes(x = lon, y = lat, group = id),
                   dat = gcircles, lwd = 0.4, alpha = 0.5)

# plot points, sized by area
p <- p + geom_point(aes(x = lon, y = lat,
                    size = sqrt(total/pi)), dat = talks_agg,
                    colour = "#555599")

# set label for legend; rescale sizes/labels for area vs. diameter
p <- p + scale_size("Total participants\n(both events)",
                    limits = c(0, max(sqrt(talks_agg$total / pi)) + 1),
                    breaks = sqrt(c(10, 50, 100) / pi),
                    labels = c(10, 50, 100), range = c(1, 10))
p <- p + theme_bw()

# gets rid of axis text/ticks, moves legend to lower left vs. right side
p <- p + theme(axis.text = element_blank(),
               axis.title = element_blank(),
               axis.ticks = element_blank(),
               panel.grid = element_blank(),
               legend.position = c(0.092, 0.15))
p
```

# Creating `.kml` files for Google Earth

Perhaps you want to work with Google Earth. `.kml` format works well and there are some links I found which helped accomplish this.

- [R/kml walkthrough](http://www.nceas.ucsb.edu/scicomp/usecases/shapeFileToKML)
- [SO question on R to kml](http://stackoverflow.com/questions/7813141/how-to-create-a-kml-file-using-r)
- [SO question on styling](http://stackoverflow.com/questions/21487010/assistance-with-name-and-styleurl-in-kml-when-using-writeogr-from-rgdal)
- [Tutorial on plotKML package](http://gsif.isric.org/doku.php?id=wiki:tutorial_plotkml))

We can take the data from earlier and re-create the great circle plot in Google Earth

```{r }
# increase the size for reasonable icon scales
talks_agg$size <- 100 * talks_agg$total

# need to turn table into a spatial object
talks_sp <- talks_agg
coordinates(talks_sp) <- c("lon", "lat")                    # converts to spatial object
proj4string(talks_sp) <- CRS("+init=epsg:4238")             # G Earth coordinate system
talks_ll <- spTransform(talks_sp,
                        CRS("+proj=longlat +datum=WGS84"))  # Something else G Earth wants

# open a file connection and write the data with custom icon/size/label
kml_open("./data/talks-w-gcircles.kml")
kml_layer.SpatialPoints(talks_ll, colour = "white", labels = city, size = total,
                        shape="http://upload.wikimedia.org/wikipedia/commons/a/af/Tux.png")
kml_close("./data/talks-w-gcircles.kml")
```

I think `plotKML` should be able to create the arcs, but I didn't figure out how... so here we're just writing the fairly simple `xml` tags.

```{r }
# for each point set, generate the following kml syntax
gcircs <- lapply(1:nrow(end), function(i) {
  paste0("<Placemark><LineString><tesselate>1</tesselate><coordinates>",
         start$lon, ",", start$lat, " ", end[i, "lon"], ",", end[i, "lat"],
         "</coordinates></LineString></Placemark>") })

# combine the results into a single table
gcircs <- do.call(rbind, gcircs)

# print out each line of the above table
# retult can simply be pasted after last <Placemark> in .kml file
write.table(gcircs, row.names = F)
```

# Plotting a trip/path

Something else you might want to do is plot a GPS path. I got some data from [AndroSensor</span>](https://play.google.com/store/apps/details?id=com.fivasim.androsensor&hl=en) on my phone via my morning trip to work and then plotted it, along with the logged speed.

```{r }
# the file from my phone
gps <- read.csv("./data/gps-data.csv", sep = ";")

# subset to columns we're interested in
gps <- gps[, c(10, 11, 13, 18)]

# give the columns sensical names
names(gps) <- c("lat", "lon", "speed", "time")

# look at the resultant table if you want
head(gps, 5)

# we can get a map using the min/max coordinates as our bounding box
box <- c(left = min(gps$lon), bottom = min(gps$lat),
         right = max(gps$lon), top = max(gps$lat))

gps_map <- get_map(location = box, source = "stamen",
                   maptype = "terrain", crop = T)

# plot as usual, specifying the dot color as mapping to the speed column in the data
p <- ggmap(gps_map)
p <- p + geom_point(aes(x = lon, y = lat, colour = speed), data = gps, size = 3)
p <- p + scale_colour_continuous(low = "black", high = "red", na.value = NA)
```

# Public transportation efficiency

You can read more about the roots of this [here](https://github.com/tcrug/public-transpo). Basically, I wanted to examine why some public transportation systems were so low/high in terms of their efficiency (energy consumption required to transport people around). I was using bubble plots, but the data lended itself to overplotting a map, so that's what we're doing here.

```{r transpo}
# read in the data
transpo <- read.csv("./data/public-transpo.csv")

transpo_agg <- ddply(transpo, .(city, state), summarize,                    # stats by state
                     btus_pmile_ave = mean(btus_pmile),
                     density = mean(population) / mean(service_area_sq_mi))

# concatenate city and state for lookups with geocode()
transpo_agg$lookup <- paste0(transpo_agg$city, ", ", transpo_agg$state)

coords <- geocode(transpo_agg$lookup)                                       # city/state string

transpo_agg <- cbind(transpo_agg, coords)                                   # get lat/lon

# once you geocode all those locations, I hightly recommend storing the results
# as a file so you don't have to re-run that part
write.table(transpo_agg, file = "./data/transpo-agg-geocoded.csv", row.names = F, sep = ",")

# read in saved data
plot <- read.csv("./data/transpo-agg-geocoded.csv")

# remove AK and HI since putting them in a plot makes the rest of the US really small
plot <- plot[plot$state != "AK" & plot$state != "HI", ]

# get our map
usa <- map_data("state")

# sort so bigger bubbles are plotted first
plot <- plot[order(plot$btus_pmile_ave), ]

# the plot, starting with US map
p <- ggplot() + geom_polygon(aes(x = long, y = lat, group = group),
                             data = usa, fill = "gray95", colour = "gray10")

# add points, colored by efficiency, sized by passengers served per square mile
p <- p + geom_point(aes(x = lon, y = lat, colour = btus_pmile_ave,
                    size = sqrt(density)/pi), data = plot)

# log scale the colors to reduce the effect of the huge spread
p <- p + scale_colour_gradient("BTUs / passenger-mile", trans = "log")

# we used the sqrt() function above to size the circles by area properly
# so we need to adjust the labels to show area instead of the calculated
# diameter we used above
p <- p + scale_size_continuous("Passengers / Sq. Mile", breaks = c(10, 20, 30, 40),
                               labels = round((c(10, 20, 30, 40)*pi)^2, 0),
                               range = c(1, 8))
# remove axis labels and ticks
p <- p + scale_x_continuous("") + scale_y_continuous("")
p <- p + theme_bw() + theme(axis.text = element_blank(), axis.ticks = element_blank())
p
```


# Converting data to `.json` for d3

If you want to use d3 or something else that likes `.json` format, you can write it pretty easily. Initially I was using [this function](http://theweiluo.wordpress.com/2011/09/30/r-to-json-for-d3-js-and-protovis/) from someone's blog, but was recently alerted to the `jsonlite` package, which is dead simple.

```{r }
# take the plot dataset from the transpo map above and convert to json
plot_json <- toJSON(plot)              # convert with jsonlite

# open a file, write the json object we created, close the file
file_con <- file("./d3/transpo.json")  # create file
writeLines(plot_json, file_con)        # write results to file
close(file_con)
```

The results look like so:

```{r eval=FALSE}
[ { "city" : "Abilene", "state" : "TX", "btus_pmile_ave" : 13574.1, "density" : 2547.39,
    "lookup" : "Abilene, TX", "lon" : -99.73, "lat" : 32.45 }, ...
```

# Visualizing work activity

I tracked my time for a while with this [Timesheet](https://play.google.com/store/apps/details?id=com.rauscha.apps.timesheet) Android app. My idea was to show the various buildings/groups I interacted with throughout the year. On the cool end of the extreme, I thought it would be neat to track things like brainstorming sessions and then be able to show the interaction/experimentation history from some key interaction all the way through to product commercialization.

Instead, I fizzled out, but here's plotting the data I *did* collect on top of 3M buildings.

```{r }
# read in data
time <- read.csv("./data/timesheet.csv")              # read data

time <- time[!is.na(time$Location), ]                 # remove entries w/ no loc

# aggregate total time per location
time <- ddply(time, .(Location), summarize,           # total time per loc
              total = sum(Duration)) 

mmm_box <- c(left = -93.0053, bottom = 44.9485,       # bounding box
             right = -92.9844, top = 44.9632)

mmm <- get_map(location = mmm_box, source = "stamen", # get map
               maptype = "toner", crop = T)

# figure out building coords and merge with timesheet data
# there's probably a better way, but I simply created a background map and used trial/error
# to get the points centered right
bldgs <- data.frame(loc = time$Location))
bldgs$lat <-  c(44.9515, 44.9576, 44.9525, 44.9545, 44.9577, 44.955,  44.9512, 44.9619, 44.9585)
bldgs$lon <- -c(93.0033, 92.9935, 92.9955, 92.9982, 92.9998, 92.9897, 93.0009, 93.00,   92.9942)

# merge building coordinates with timesheet data
time <- merge(time, bldgs, by.x = "Location", by.y = "loc")

# line from my main building to each of the other ones
lines <- data.frame(lat = c(rep(bldgs[2, "lat"], 8), bldgs[c(1, 3:9), "lat"]),
                    lon = c(rep(bldgs[2, "lon"], 8), bldgs[c(1, 3:9), "lon"]),
                    grp = rep(1:8, 2))

# bacground map
p <- ggmap(mmm)

# lines
p <- p + geom_line(aes(x = lon, y = lat, group = grp),
                   data = lines, colour = "steelblue")

# points sized by total hours per building
p <- p + geom_point(aes(x = lon, y = lat, size = sqrt(total)/pi),
                    data = time, fill = "steelblue", colour = "black", pch = 21)

# set sensible breaks and size ranges for hours
p <- p + scale_size_continuous("Hours", breaks = c(0.5, 1, 1.5, 2),
                               labels = c(2.5, 10, 20, 40), range = c(2, 6))

# ditch axis labels/ticks
p <- p + scale_x_continuous("") + scale_y_continuous("")
p <- p + theme_bw() + theme(axis.text = element_blank(), axis.ticks = element_blank())
p

```

# References

-   [ggmap: Spatial Visualization with ggplot2](http://stat405.had.co.nz/ggmap.pdf), Kahle \\& Wickham
-   [How to draw good looking maps in R](http://uchicagoconsulting.wordpress.com/2011/04/18/how-to-draw-good-looking-maps-in-r/), uchicagoconsulting
-   [Great circles on a recentered worldmap](http://www.stanford.edu/~cengel/cgi-bin/anthrospace/great-circles-on-a-recentered-worldmap-in-ggplot), AnthroSpace
-   [How to map connections with great circles](http://flowingdata.com/2011/05/11/how-to-map-connections-with-great-circles/), FlowingData
-   [SO question on creating `.kml` files](http://stackoverflow.com/questions/21487010/assistance-with-name-and-styleurl-in-kml-when-using-writeogr-from-rgdal)
-   For any given package, find the documentation on [CRAN](http://cran.r-project.org/)

\vspace{0.5cm}

Code and files from this presentation are on [github](https://github.com/jwhendy/devFest-geo)!